Type theories as quotient inductive-recursive types

### 2025-02-07 tex/main.tex
Fix some typos in S4.2

\begin{abstract}
  Defining type theory in type theory is a long-standing challenge and a benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
\end{abstract}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce t'$ for $t \equiv t'$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\LT{We may wonder if the last three identities can be made strict.
Can we find a complete set of rules?}

@@

\begin{remark}
To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ \emph{already} requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
\end{remark}

@@

\subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
Our definition has the standard $\Set$-theoretic model, justifying the soundness of our definition, given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{alignat*}{5}
  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
\end{alignat*}
with the additional methods to its QII counterpart:
\begin{align*}
  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M} (\sigma^M ; t^M)  \\
                         & : (\Gamma^M \to \Delta^M) \to ((\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
\end{align*}
where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.

@@

In contrast, in our QIIR definition, many identities just hold strictly.
Consider
\begin{alignat*}{5}
  [\idS]^M                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
\end{alignat*}
where $[\idS]^M$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$, strictly equal to another strict equality $(\lam{\gamma}(\pi_1\gamma, \pi_2\gamma)) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.

The additional rules we have used to enable further reduction are derivable already, and they can be justified by the same standard model with simpler proofs.
Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
\section{Formal implementation} \label{sec:formal-implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

@@

\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.}

>>>>

\begin{abstract}
  Defining type theory in type theory is a long-standing benchmark for the strength and expressivity of type theory and a proof assistant.
  Altenkirch and Kaposi used quotient inductive-inductive types to elegantly represent exactly the well-typed terms of type theory, with the price to be paid that terms frequently contain transports along equality constructors.
  As a result, proofs need to painstakingly account for coercions alongside the interesting part.
  This %mess
  has been dubbed `transport hell', but is typically suppressed in informal presentation for clarity, hiding the gap between the intention and the formalisation of type theory in type theory.
  In this paper, we aim to shorten this gap, using quotient inductive-inductive-recursive types and definitions by overlapping patterns, reducing the use of transports.
  As a case study, we investigate type theory with a universe and $\Pi$-types, and compare our definition with quotient inductive-inductive definitions.
\end{abstract}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & \defeq \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & \defeq A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & \defeq [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
However, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}.
By design, definitions by pattern matching will be elaborated to a form that corresponds to the \emph{eliminator} either using \emph{first-match semantics}~\cite{Cockx2020a} for function clauses or an explicit construct for pattern matching, so that \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will only fire if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\LT{We may wonder if the last three identities can be made strict.}

@@

\begin{remark}
To state these methods, we cannot avoid using transports in identities that would need transports in its QII definition of a type theory (e.g.\ $[\idS]_{\Tm}^M$) since identities (e.g.\ $[\idS]_{\Ty}^M$) are propositional just like equality constructors in its QII definition.
However, in the elimination principle for the QII definition, the situation is even worse: the above identity $[\idS]_{\Ty}^M$ already requires $\alert{\transfib{\Ty^M\;\Gamma^M}{[\idS]_{\Ty}}{\color{black}[\idS^M]_{\Ty}^M\,A^M}}$ to coerce its type $\Ty^M\;\Gamma^M\;([\idS]\,A)$ to $\Ty^M\;\Gamma^M\;A$ to match the index $A$ of $A^M$, which are not needed in our QIIR definition.
That is, the eliminator for the QIIR definition requires fewer transports than its QII counterpart.
\end{remark}

@@

\subsection{The \texorpdfstring{$\Set$}{Set}-theoretic model} \label{subsec:std-model}
The standard $\Set$-theoretic model for our definition, justifying its soundness, is given by the same motives as those for its QII counterpart~\cite[Section~4]{Altenkirch2016a}:
\begin{alignat*}{5}
  \Ctx^M & \defeq \lam{\nameless} \; && \Set & \qquad \Sub^M & \defeq \lam{\Gamma^M\,\Delta^M\,\nameless}\; && \Gamma^M \to \Delta^M \\
  \Ty^M  & \defeq \lam{\Gamma^M\, \nameless} && \Gamma^M \to \Set & 
  \Tm^M  & \defeq \lam{\Gamma^M\,A^M\,\nameless} && (\gamma : \Gamma^M) \to A^M\,\gamma
\end{alignat*}
with the additional methods to its QII counterpart:
\begin{align*}
  \blank\uparrow^M\blank & \defeq \lam{\sigma^M\,A^M\,\gamma} (\sigma^M\,(\pi_1\,\gamma) , \pi_2\,\gamma) & \\
                         & : (\sigma^M : \Gamma^M \to \Delta^M) (A^M : \Delta^M \to \Set) \to \Sigma\;\Gamma^M\;(\lam{\gamma}A^M\,(\sigma^M\,\gamma)) \to \Sigma\,\Delta^M\,A^M \\
  [\blank]_{\Tm}^M\blank & \defeq \lam{\sigma^M\,t^M\,\gamma} (t^M\,(\sigma^M\,\gamma))  \\
                         & : (\sigma^M : \Gamma^M \to \Delta^M) (t^M : (\delta : \Delta^M) \to A^M\,\delta) \to (\gamma : \Gamma^M) \to A^M\,(\sigma^M\,\gamma)
\end{align*}
where $[\blank]_{\Tm}^M\blank$ is the same as the method for the explicit term substitution $[\blank]^M\blank$.

@@

In contrast, in our QIIR definition, many identities just hold strictly.
Consider
\begin{alignat*}{5}
  [\idS]^M                   & : [ \idS^M ]^M_{\Ty}\, A^M && = A^M \\
  \idS{\mathord{\uparrow}}^M & : \idS^M \uparrow^M A^M    && =^{\lam{B^M}\Sub^M\;(\Gamma^M, A^M)\;\Delta^M\;(\idS \uparrow A)}_{\alert{[\idS]^M}} \idS^M
\end{alignat*}
where $[\idS]^M$ does not need a transport and is just an strict equality $(\lam{\gamma} A^M\,\gamma) \judgeq A^M$ in the model by the $\eta$-law for $\Pi$-types.
Hence, $\idS{\mathord{\uparrow}}^M$ becomes $(\idS^M \uparrow^M A^M) = \idS^M$ and by definition is a strict equality $\lam{\gamma}(\pi_1\gamma, \pi_2\gamma) \judgeq \lam{\gamma}\gamma$ by the $\eta$-law for $\Sigma$-types.

The additional rules we have used to enable further reduction are derivable in type theory, and they can be justified by the same standard model with simpler proofs.
Now is nature to ask: what is the theoretical difference between the QIIR and QIIR definitions?
\subsection{Comparison with type theory as QIIT} \label{subsec:equivalence}
%\subsubsection{Syntactic equivalence}
%\subsubsection{Semantics equivalence}

\section{Formal implementation} \label{sec:formal-implementation}
\LT{Mention the NbE for SC here}
\LT{Compare some formal proofs with pen-and-paper proofs here}
\section{Discussion}
\subsection{Related work}
\paragraph*{Formalisation of type theory in type theory}
\cite{Danielsson2006,Altenkirch2016a,Chapman2009}
\cite{Altenkirch2017}
\cite{Munoz1998}
\cite{Dybjer1996,Castellan2021}
\LT{Dybjer \cite{Dybjer1996} already says cwf is inductive-recursive}

@@

\LT{OTT also has a natural setoid model (which should be useful to manifest the difference between recursion and equality constructor) but quotient (inductive-)inductive(-recursive) types in OTT are underdeveloped.
It will be nice to have OTT equipped with QII(R)T on a par with HIIT (or Cubical Agda).}
